---
title: "Covid data EDA"
output: html_notebook
---

```{r}
library(tidyverse)
covid_cases_deaths <- read_csv("https://raw.githubusercontent.com/36-SURE/36-SURE.github.io/main/data/covid_cases_deaths.csv")
```

Hypothesis 1: More densely populated counties have higher rates of covid cases. 

Join main df with data on county area

Source: https://www.rural.pa.gov/download.cfm?file=Resources/excel/CountyProfile.xlsx

```{r}
pop_density_df <- read_csv("https://raw.githubusercontent.com/LJxse27/Group-5/main/CountyProfile.csv")

county_area <- pop_density_df |> 
  filter(`...1` == 'Square Land Miles, 2020') |> 
  select(-Pennsylvania, -'...1') |> 
  pivot_longer(cols = everything(), names_to = "county", values_to = 'Area') |> 
  mutate(Area = as.numeric(gsub(",", "", Area))) |> 
  mutate(county = stringr::str_remove(county, " County"))

covid_cases_deaths_area <- covid_cases_deaths |> 
  left_join(county_area, by = "county")
```

## Visualize relationship between pop density, cases rate, and deaths rate


```{r}
# Create pop density variable
covid_with_density <- covid_cases_deaths_area |> 
  group_by(county) |> 
  mutate(pop_density = population / Area)
```

Factor covid density into 3 categories by percentile

```{r}
# Find 33 and 66th quartlies
percentiles = quantile(covid_with_density$pop_density, probs = c(0.33, 0.66))
percentiles
```

Plot variables using violin plot and boxplots, excluding outliers

```{r}
# Identifying the outliers
Q1_cases = quantile(covid_with_density$cases_rate, 0.25)
Q3_cases = quantile(covid_with_density$cases_rate, 0.75)
IQR_cases = Q3_cases - Q1_cases

Q1_deaths = quantile(covid_with_density$deaths_rate, 0.25)
Q3_deaths = quantile(covid_with_density$deaths_rate, 0.75)
IQR_deaths = Q3_deaths - Q1_deaths

# Remove outliers
lower_bound_cases = Q1_cases - 1.5 * IQR_cases
upper_bound_cases = Q3_cases + 1.5 * IQR_cases

lower_bound_deaths = Q1_deaths - 1.5 * IQR_deaths
upper_bound_deaths = Q3_deaths + 1.5 * IQR_deaths

filtered_data <- covid_with_density |> 
  group_by(county) |> 
  filter(lower_bound_cases <= cases_rate & cases_rate <= upper_bound_cases & lower_bound_deaths <= deaths_rate & deaths_rate <= upper_bound_deaths) |> 
  mutate(density_cat = cut(pop_density, 
      breaks = c(-Inf, percentiles[1], percentiles[2], Inf),
      labels = c("Low Density", "Medium Density", "High Density")))

# Graph cases data without outliers
cases_by_density <- filtered_data |> 
  ggplot(aes(x = cases_rate, y = density_cat)) +
  geom_boxplot(alpha = 0.2, width = 0.1, aes(fill = density_cat)) +
  geom_violin(alpha = 0.5, aes(fill = density_cat)) +
  labs(title = "COVID cases by pop. density",
       x = "Cases rate",
       y = "Population Density",
       fill = "Populaiton Density") +
  theme(legend.position = "none")
```


```{r}
# Graph data without outliers
deaths_by_density <- filtered_data |> 
  ggplot(aes(x = deaths_rate, y = density_cat)) +
  geom_boxplot(alpha = 0.2, width = 0.1, aes(fill = density_cat)) +
  geom_violin(alpha = 0.5, aes(fill = density_cat)) +
  labs(title = "COVID deaths by pop. density",
       x = "Deaths rate",
       y = "",
       fill = "Population Density") +
  theme(legend.position = "none")
```


```{r}
library(cowplot)
plot_grid(cases_by_density, deaths_by_density)
```

Hypothesis 2: counties with higher cumulative cases (cases_cume_rate) have lower new case rate (cases_rate) due to herd immunity effects 

```{r}
covid_cases_deaths |> 
  filter(county %in% c("Allegheny", "Montgomery", "Philadelphia")) |>
  mutate(date = as.Date(date)) |> 
  filter(date >= '2020-10-01') |> 
  ggplot(aes(x = cases_cume_rate, y = cases_rate)) +
  geom_point(size = 0.5, alpha = 0.3) +
  geom_smooth(method = "loess") +
  facet_wrap(~county, scales = "free") +
  labs(title = "Covid cases rate by cumulative cases rate",
       x = "Cumulative cases rate",
       y = "New cases rate")
```
Potential explanations for why the curves are decreasing: 
- spike after thanksgiving, people won't test for covid they got over christmas until January


## Clustering

Questions:
- counties with early peaks vs late peaks (max cases date)?
- which areas had most severe outbreaks (max cases rate)?

```{r}
library(dplyr)

covid_agg <- covid_cases_deaths |> 
  group_by(county) |> 
  summarize(max_cases_rate = max(cases_rate),
         max_cases_date = date[which.max(cases_rate)],
         median_cases = median(cases_rate))
```


```{r}
library(lubridate)
# Convert max_cases_date to numeric (e.g., number of days since a reference date)
covid_agg <- covid_agg |> 
  mutate(max_cases_date_numeric = as.numeric(difftime(max_cases_date, as.Date("2020-04-01"), units = "days")))
```


```{r}
# Scale the data for clustering
covid_scaled <- covid_agg |>
  select(max_cases_rate, max_cases_date_numeric, median_cases) |>
  scale()
```

```{r}
# Create distance matrix
distance_matrix <- dist(covid_scaled)
```

```{r}
hc_complete <- hclust(distance_matrix, method = "complete")
```


```{r}
# Convert to dendrogram for plotting
dend <- as.dendrogram(hc_complete)

# Convert dendrogram to data frame for ggplot2
dend_data <- dendro_data(dend)

# Create a data frame with the labels
label_data <- dend_data$labels %>%
  mutate(label = covid_agg$county[order.dendrogram(dend)])
# Plot the dendrogram with labels
ggplot() +
  geom_segment(data = dend_data$segments, 
               aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_text(data = label_data, 
            aes(x = x, y = y, label = label), 
            hjust = -0.1, size = 2) + # Adjust size as needed
  coord_flip() + 
  scale_y_reverse() + 
  geom_hline(yintercept = 4.5, color = "red") +
  labs(y = "Dissimilarity between clusters", 
       x = NULL,
       title = "Dendrogram of County Clustering Based on COVID-19 Metrics") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

```{r}
# Cut dendrogram into clusters
num_clusters <- 4
clusters <- cutree(hc_complete, k = num_clusters)
```

```{r}
# Create data frame with counties and their cluster assignments
cluster_assignments <- data.frame(county = covid_agg$county, cluster = as.factor(clusters))
```


```{r}
# Merge the two datasets
covid_agg_with_clusters <- left_join(covid_agg, cluster_assignments, by = "county")
```


```{r}
library(sf)
library(tigris)
# Get the shapefile for Pennsylvania counties
pa_counties <- counties(state = "PA", cb = TRUE, class = "sf")
```
```{r}
# Merge geographic data with cluster data
pa_counties <- pa_counties |> 
  left_join(covid_agg_with_clusters, by = c("NAME" = "county"))
```


```{r}
# Plot clusters on choropleth
pa_counties |> 
  ggplot() +
  geom_sf(aes(fill = cluster)) +
  # scale_fill_viridis_d(name = "Cluster", option = "inferno", na.value = "grey50") +
  geom_sf_text(aes(label = NAME), size = 2, check_overlap = TRUE) +
  labs(title = "Choropleth Map of Pennsylvania Counties by Cluster",
       subtitle = "Clusters based on COVID-19 Cases Data",
       caption = "Source: https://raw.githubusercontent.com/36-SURE/36-SURE.github.io/main/data/covid_cases_deaths.csv") +
  scale_fill_brewer(palette = "Set2", na.value = "grey50") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
  
```

Summary statistics for each cluster:

```{r}
cluster_summary <- covid_agg_with_clusters |> 
  group_by(cluster) |> 
  summarize(
    median_max_cases_rate = median(max_cases_rate, na.rm = TRUE),
    median_max_cases_date_numeric = median(max_cases_date, na.rm = TRUE),
    median_median_cases = median(median_cases, na.rm = TRUE)
  )
```



```{r}
# Boxplots for max cases rates by cluster
covid_agg_with_clusters |> 
  ggplot(aes(x = factor(cluster), y = max_cases_rate)) +
  geom_boxplot(alpha = 0.8, aes(fill = factor(cluster))) +
  labs(title = "Distribution of max cases rate by cluster",
       fill = "Cluster",
       x = "Cluster", 
       y = "Max cases rate") +
  scale_fill_brewer(palette = "Set2", na.value = "grey50") +
  theme_light() 
  
```

```{r}
# Boxplots for max cases dates by cluster
covid_agg_with_clusters |> 
  ggplot(aes(x = factor(cluster), y = max_cases_date)) +
  geom_boxplot(alpha = 0.8, aes(fill = factor(cluster))) +
  labs(title = "Distribution of max cases date by cluster",
       fill = "Cluster",
       x = "Cluster",
       y = "Max cases date") +
  scale_fill_brewer(palette = "Set2", na.value = "grey50") +
  theme_light() 
  
```

```{r}
# Boxplots for median cases rates by cluster
covid_agg_with_clusters |> 
  ggplot(aes(x = factor(cluster), y = median_cases)) +
  geom_boxplot(alpha = 0.8, aes(fill = factor(cluster))) +
  labs(title = "Distribution of median cases rate by cluster",
       fill = "Cluster",
       x = "Cluster",
       y = "Median cases rate") +
  scale_fill_brewer(palette = "Set2", na.value = "grey50") +
  theme_light() 
  
```

Visualize max cases in Cameron and Crawford

```{r}
covid_cases_deaths |> 
  filter(county %in% c("Crawford", "Cameron")) |> 
  ggplot(aes(x = date, y = cases_rate)) +
  facet_wrap(~county) +
  labs(title = "Visualizing cases spikes in Cameron and Crawford",
       x = "Date", 
       y = "Cases rate") +
  geom_line()
```

Visualize Centre cases rate

```{r}
covid_cases_deaths |> 
  filter(county == "Centre") |> 
  ggplot(aes(x = date, y = cases_rate)) +
  facet_wrap(~county) +
  labs(title = "Visualizing date of covid spikes in Centre County",
       x = "Date",
       y = "Cases rate") +
  geom_line()
```








