---
title: "R Notebook"
output: html_notebook
---

Source: https://www.rural.pa.gov/download.cfm?file=Resources/excel/CountyProfile.xlsx


```{r}
library(tidyverse)
covid_cases_deaths <- read_csv("https://raw.githubusercontent.com/36-SURE/36-SURE.github.io/main/data/covid_cases_deaths.csv")
```

List of hypotheses:
* higher population density is associated with higher covid case rates
* counties in higher latitudes have different covid case rates than those at lower latitudes
* the seven day rolling average of new cases (cases_avg_new_rate) shows distinct trends over time related to policy change
* countries with higher cumulative cases (cases_cume_rate) have lower new case rate (cases_rate) due to herd immunity effects 


Hypothesis: More densely populated counties have higher rates of covid cases. 

Variables: 
* pop_density_cat (fct)
* cases_rate (dbl)

Graph options:
* bar chart
* side by side box-plots
* violin plots
* ridgeline plot

## Join df with data on county area

```{r}
pop_density_df <- readxl::read_xlsx("/Users/lissandroalvarado/Downloads/ACSST5Y2020.S1902-2024-06-10T184913.xlsx", sheet = 2)

county_area <- pop_density_df |> 
  rename(variable_label = ...1) |>
  filter(variable_label == 'With wages or salary income') |>
  #select(-Pennsylvania, -'...1') |> 
  pivot_longer(cols = -variable_label, names_to = "county", values_to = 'Household Income') |>
  janitor::clean_names() |> 
  mutate(household_income = as.numeric(gsub(",","", household_income))) |>
  mutate(county = stringr::str_remove(county, " County, Pennsylvania")) |> select(-variable_label)
```

```{r}
#covid_cases_deaths_area <- covid_cases_deaths |> 
 # left_join(county_area, by = "county")
head(pop_density_df)
```


## Visualize relationship between pop density and cases


```{r}
# Create pop density variable
covid_with_density <- covid_cases_deaths_area |> 
  group_by(county) |> 
  mutate(pop_density = population / Area)
```


```{r}
# Bar chart
library(dplyr)
library(forcats)
covid_with_density |> 
  ggplot(aes(x = cases_rate, y = reorder(county, cases_rate))) +
  geom_col(fill = "midnightblue") +
  labs(title = "COVID case rate by ennsylvania county",
       x = "Cases rate",
       y = "County")
```

Factor covid density into 3 categories by percentile


```{r}
# Find 33 and 66th quartlies
percentiles = quantile(covid_with_density$pop_density, probs = c(0.33, 0.66))
percentiles
```

Create violin plots with the 3 categories

```{r}
covid_with_density <- covid_with_density |> 
  mutate(density_cat = cut(pop_density, 
      breaks = c(-Inf, percentiles[1], percentiles[2], Inf),
      labels = c("Low Density", "Medium Density", "High Density")))
```

```{r}
covid_with_density |> 
  ggplot(aes(x = cases_rate, y = density_cat)) +
  geom_boxplot(alpha = 0.5, width = 0.2, aes(fill = density_cat)) +
  geom_violin(alpha = 0.5, aes(fill = density_cat))
```

There appears to be outliers.

```{r}
# Identifying the outlier
Q1 = quantile(covid_with_density$cases_rate, 0.25)
Q3 = quantile(covid_with_density$cases_rate, 0.75)
IQR = Q3 - Q1

# Remove outlier
lower_bound = Q1 - 1.5 * IQR
upper_bound = Q3 + 1.5 * IQR

filtered_data <- covid_with_density |> 
  group_by(county) |> 
  filter(lower_bound <= cases_rate & cases_rate <= upper_bound) |> 
  mutate(density_cat = cut(pop_density, 
      breaks = c(-Inf, percentiles[1], percentiles[2], Inf),
      labels = c("Low Density", "Medium Density", "High Density")))

# Graph data without outliers
filtered_data |> 
  ggplot(aes(x = cases_rate, y = density_cat)) +
  geom_boxplot(alpha = 0.2, width = 0.1, aes(fill = density_cat)) +
  geom_violin(alpha = 0.5, aes(fill = density_cat)) +
  labs(title = "COVID cases by population density, excluding outliers",
       x = "Cases rate",
       y = "Population Density",
       fill = "Populaiton Density") +
  theme(legend.position = "none")
```

Which counties are low density and what is their death rate?

```{r}
filtered_data |> 
  group_by(county) |> 
  filter(density_cat == "Low Density") |> 
  distinct(county, .keep_all = TRUE)
```

Ridgeline plot

```{r}
library(ggridges)
filtered_data |> 
  ggplot(aes(x = cases_rate, y = density_cat)) +
  geom_density_ridges(scale = 0.8, alpha = 0.5, aes(fill = density_cat)) +
  theme(legend.position = "none") +
  labs(title = "COVID cases by population density, excluding outliers",
     x = "Cases rate",
     y = "Population Density",
     fill = "Populaiton Density")
```




```{r}
names(covid_cases_deaths)
```

```{r}
glimpse(covid_cases_deaths)
```

```{r}
summary(covid_cases_deaths$deaths_rate)
```

```{r}
summary(covid_cases_deaths$date)
```


Potential questions:
- how do covid deaths vary overtime by county (Allegheny vs Philly)
- map of pennsylvania and density of cases
- compare cases vs deaths based on region
- population overtime and covid deaths
- does the population density of counties affect covid cases / deaths?


Question: How do covid deaths vary overtime by county (comparing 3 major counties)

```{r}
library(zoo)
covid_cases_deaths2 <- covid_cases_deaths |> 
  filter(county %in% c("Allegheny", "Philadelphia", "Montgomery")) |> 
  group_by(county) |> 
  arrange(date) |> 
  mutate(rolling_avg_deaths = rollmean(deaths, k = 14, fill = NA, align = "right"))
```



```{r}
library(zoo)
covid_cases_deaths2 |> 
  filter(county %in% c("Allegheny", "Philadelphia", "Montgomery")) |> 
  ggplot(aes(x = date, y = deaths)) +
  geom_line(linewidth = 0.4) +
  # geom_line(aes(y = rolling_avg_deaths), color = "red") +
  facet_wrap(~ county) +
  theme_light() +
  labs(title = "Covid deaths of major counties over time",
       x = "Date",
       y = "Covid Deaths") 
  # theme(strip.background = element_blank(),
  #       plot.title = element_text(hjust = 0.5, 
  #                                 face = "bold"))
```

```{r}
covid_cases_deaths3 <- covid_cases_deaths |> 
  filter(county %in% c("Allegheny", "Philadelphia")) |> 
  group_by(county) |> 
  arrange(date) |> 
  mutate(rolling_avg_death_rate = rollmean(deaths_rate, k = 14, fill = NA, align = "center"))
```


By proportion

```{r}
covid_cases_deaths3 |> 
  filter(county %in% c("Allegheny", "Philadelphia")) |> 
  ggplot(aes(x = date, y = deaths_rate)) + 
  geom_line(linewidth=0.5, alpha=0.8) +
  geom_line(aes(y = rolling_avg_death_rate), color = "red") +
  facet_wrap(~county) +
  theme_light()
  
```

How long it takes for cases to turn into deaths - time series with cases and deaths


Hypothesis: countries with higher cumulative cases (cases_cume_rate) have lower new case rate (cases_rate) due to herd immunity effects 

```{r}
covid_cases_deaths |> 
  mutate(date = as.Date(date)) |> 
  filter(date >= '2020-12-05') |> 
  ggplot(aes(x = cases_cume_rate, y = cases_rate)) +
  geom_point(size = 0.5, alpha = 0.3) +
  geom_smooth(method = "loess") +
  labs(title = "Covid cases rate by cumulative cases rate",
       x = "Cumulative cases rate",
       y = "New cases rate")
```






















